Supporting Information:\\Black-swan events in animal populations

### 2016-05-27 text/ms.tex
Edits

We fit population dynamics models in which the process noise is
drawn from a Student-t distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
distributions.
% For example, at \(\nu = 2\), an event
% more than 4 standard deviations below the mean is expected
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a t distribution with \(\nu = 2\)
compared to once every 32,000 years with a normal distribution
% more than four standard deviations below the mean is almost 70,000 times higher
% than the probability of drawing such a value from a normal distribution
% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

@@

We selected abundance time series from the Global Population Dynamics
Database \citep{gpdd2010} (GPDD), which contains nearly 5,000 time series of
abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
filtered the data (Supporting Information) to remove populations from less
reliable data sources, and those without sufficient data for our models, and
then interpolated some missing values (sensu Brook \textit{et
al.} \citep{brook2006a}). Our interpolation affected only \(\sim\)
\interpPointsPerc \% of the final data points (Table~S1) and none of the data
points that were later considered black-swan events. Our final dataset
contained \NPops\ populations across \NOrders\ taxonomic orders and seven
taxonomic classes, with a median of \medianTimeSteps\ time steps (range
\minTimeSteps--\maxTimeSteps) (Table~S1).

@@

For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by Gelman \emph{et al.} \citep{gelman2008d} for intercepts of
regression models. Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et
al.} \citep{gelman2006c} and Gelman \emph{et al.} \citep{gelman2008d} and the
specific scale parameter of \(2.5\) is based on our expected range of the value
in nature from previous studies \citep{connors2014}. In our testing of a
subsample of populations, our parameter estimates were not qualitatively
changed by switching to a uniform prior on \(\sigma\), but the weakly
informative Half-Cauchy prior sped up chain convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by Fernandez and Steel \citep{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid drifting towards infinity. In
any case, for \(\nu > 20\) the t distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the t distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the
population as heavy tailed.

We fit our models with Stan 2.6.0 \citep{stan-manual2015, hoffman2014}, and R
3.2.0 \citep{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again. These thresholds are in
excess of the minimums recommended by Gelman \emph{et al.} \citep{gelman2006a} of
\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
confidence intervals. In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
total samples) by which all chains were deemed to have sufficiently converged.
The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
requires fewer iterations to obtain equivalent effective sample sizes than the
Gibbs or Metropolis--Hastings algorithms \citep{stan-manual2015}.

Specifically, we fit population dynamics models in which the process noise is
drawn from a Student-t distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
Fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
distributions. For example, at \(\nu = 2\), the probability of drawing a value
more than five standard deviations below the mean is almost 70,000 times higher
than the probability of drawing such a value from a normal distribution
(\(0.02\) vs.\ \(2.9\cdot10^{-7}\)). As \(\nu\) approaches infinity, the
distribution approaches the normal distribution (Fig.~1).

@@

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are mostly concerned about when categorizing
populations as heavy tailed. We furthermore considered a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in Gelman \emph{et al.} \citep{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here (not shown).

@@

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a t distribution with heavy
tails \citep{allen2001}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \citep{schindler2010}, or an intrinsic change in population variability
through time \citep{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \citep{nunez2012} including extreme climate
events \citep{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black swans, or sharp changes in human
pressures such hunting, fishing, or habitat destruction. Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \citep{kirby2009} or a rare alignment of
non-synergistic forces \citep{denny2009}.

>>>>

We fit population dynamics models in which the process noise is
drawn from a Student-t distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
Fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
distributions.
% For example, at \(\nu = 2\), an event
% more than 4 standard deviations below the mean is expected
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a t distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution
% more than four standard deviations below the mean is almost 70,000 times higher
% than the probability of drawing such a value from a normal distribution
% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

@@

We selected abundance time series from the Global Population Dynamics
Database \citep{gpdd2010} (GPDD), which contains nearly 5,000 time series of
abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
filtered the data (Supporting Information) to remove populations from less
reliable data sources, and those without sufficient data for our models, and
then interpolated some missing values (sensu \citet{brook2006a}). Our interpolation affected only \(\sim\)
\interpPointsPerc \% of the final data points (Table~S1) and none of the data
points that were later considered black-swan events. Our final dataset
contained \NPops\ populations across \NOrders\ taxonomic orders and seven
taxonomic classes, with a median of \medianTimeSteps\ time steps (range
\minTimeSteps--\maxTimeSteps) (Table~S1).

@@

For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our Half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain
convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by \citet{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid drifting towards infinity. In
any case, for \(\nu > 20\) the t distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the t distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the
population as heavy tailed.

We fit our models with Stan 2.6.0 \citep{stan-manual2015, hoffman2014}, and R
3.2.0 \citep{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again. These thresholds are in
excess of the minimums recommended by \citet{gelman2006a} of
\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
confidence intervals. In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
total samples) by which all chains were deemed to have sufficiently converged.
The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
requires fewer iterations to obtain equivalent effective sample sizes than the
Gibbs or Metropolis--Hastings algorithms \citep{stan-manual2015}.

@@

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are mostly concerned about when categorizing
populations as heavy tailed. We furthermore considered a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here (not shown).

@@

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a t distribution with heavy
tails \citep{gelman2014}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \citep{schindler2010}, or an intrinsic change in population variability
through time \citep{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \citep{nunez2012} including extreme climate
events \citep{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black swans, or sharp changes in human
pressures such hunting, fishing, or habitat destruction. Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \citep{kirby2009} or a rare alignment of
non-synergistic forces \citep{denny2009}.
